Mechanising Hylomorphisms for Extraction of Executable Code

### 2025-03-24 paper/paper.tex
Containers

\subsection{Mechanising Extractable Container Functors}
\label{sec:containers}
\todo{FILL OUT}{}
The type $\forall s : S.\; P\;s \to X$ would force us to add
the functional extensionality axiom. Thus, we require that all the types used in
our definitions are \emph{setoids}, where we use \coq{x =e y} to represent
setoid equality. All morphisms need to be \emph{proper} morphisms that respect
setoid equalities. In our development, we use
\begin{coqcode}
Notation "A ~> B" := (* proper morphisms  s.t. forall x y, x =e y -> f x =e f y *).
\end{coqcode}
We add an automatic coercion from \coq{A ~> B} to \coq{A -> B} and provide
tactics to automatically discharge proofs that morphisms are proper,  for some
simple cases.

%% WARNING: removing the point about axiom K leads to leaving the use of
%% decidable validity predicates unexplained -- do not remove!
We define containers in terms of shapes, \emph{base} positions
(\coq{APos : Set}), and \emph{decidable validity predicates} that specify when a
base position is valid in a shape. Since validity predicates are in \coq{Prop},
they will be erased during extraction, and since they are decidable, they
do not require axiom K to deal with heterogeneous equalities.
Container extensions are defined as
\begin{coqcode}
Record App F X := {shape : Shape F; cont : {p : APos F | valid s p = true} -> X}.
\end{coqcode}

%% Make sure you do not cut text so much that it ends up being inconsistent with
%% the code! I just restored an old explanation and cut some of it in a consistent way
We define the setoid equality of two container extensions \coq{x y : App F X} as
\begin{coqcode}
shape x =e shape y /\ (forall p p', projT1 p = projT1 p' -> cont x p =e cont y p')
\end{coqcode}

>>>>

\subsection{Mechanising Extractable Container Functors}
\label{sec:containers}

A direct encoding of containers as presented in
Section~\ref{sec:category-theory} requires the use of the functional
extensionality axiom, and axiom K to deal with equalities of objects of type
$P(s) \to X$. Furthermore, extraction will need to insert unsafe casts for
anything of type $P(s)$.  To avoid these problems, we use \emph{setoids}, and
encode families of positions using \emph{decidable validity predicates}.

In our formalisation, we require that every type is a setoid, where \coq{x =e y}
denotes setoid equality. All morphisms need to be \emph{proper} morphisms that
respect setoid equalities. We use special notation for such morphisms, \coq{A ~>
  B}, and we add an implicit coercion from \coq{A ~> B} to \coq{A -> B}.  We
also provide tactics to automatically discharge proofs that morphisms are
proper,  for some simple cases.

%% WARNING: removing the point about axiom K leads to leaving the use of
%% decidable validity predicates unexplained -- do not remove!
We define containers in terms of shapes, \emph{base} positions (\coq{APos :
  Set}), and \emph{decidable validity predicates} that specify when a base
position is valid in a shape. Since validity predicates are in \coq{Prop}, they
will be erased during extraction, and since they are decidable, they do not
require axiom K to deal with heterogeneous equalities.  Container extensions are
defined as
\begin{coqcode}
Record App F X := {shape : Shape F; cont : {p : APos F | valid s p = true} -> X}.
\end{coqcode}
%% Make sure you do not cut text so much that it ends up being inconsistent with
%% the code! I just restored an old explanation and cut some of it in a consistent way
We define the setoid equality of two container extensions \coq{x y : App F X} as
\begin{coqcode}
shape x =e shape y /\ (forall p p', projT1 p = projT1 p' -> cont x p =e cont y p')
\end{coqcode}
